### 2024-11-20 teaching/math240f24/draft.tex
Lecture notes 19 nov

\vskip48pt

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
graphs}{74}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 15.\kern .5em Eulerian
trails and circuits}{81}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}

@@

\medskip\boldlabel\datestamp{10}{ix} Conditional and biconditional. We now examine the {\it conditional}
logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\To q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 1 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}.
$$
Within the conditional statement, $p$ is called the {\it antecedent}; this is the assumption. The
statement that is asserted, conditional that the antecendent holds, is called the {\it consequent} $q$.
You can quickly check that the relation $p\To q$ is equivalent to $\neg p \vee q$. This fact is useful
when performing mechanical simplifications.
In English, the statement ``if $p$ then $q$'' asserts a causal relation between $p$ and $q$.
Take a moment and reconcile this idea with the truth table above.
It is normal to get a little tripped up if it's your first time seeing this table.
In the first row, $q$ doesn't even happen, so it might feel weird to say that $p$
has ``caused'' $q$ to happen in this case, and
in the second row, $p$ is not true
and $q$ is true, so it might seem odd that we have set $p\To q$ to true, because there seems to be no relation
between $p$ and $q$.
But it should make sense if you think
of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
suppose your friend says, If it snows tomorrow I'll work in Trottier with you. If it doesn't snow
and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
snow and she shows up, then she still hasn't broken her promise. The only way she can break her promise is if
it snows and she doesn't come to Trottier; this situation corresponds to the only $0$-row in the truth table.

@@

\proclaim Proposition \advthm. Let $X$ be a finite nonempty set.
Let $E$ be the set of all subsets of $X$ with
even cardinality, and let $D$ be the set of all subsets of $X$ with odd cardinality.
Then $|E| = |D|$.

@@

\parenproclaim Theorem {\advthm} (Prime number theorem). For $x\in \RR$, let
$$\pi(x) = \bigl| \{ p \le x : p\ \hbox{prime}\} \bigr|.$$
Then $\pi(x) \sim x/\ln x$, in the sense that
$$\lim_{x\to\infty} {\pi(x)\over x/\ln x} = 1.$$

@@

\noindent\datestamp{22}{} \hskip\parindent
Not only are all graphs that have $K_5$ or $K_{3,3}$ as a minor nonplanar, it turns out
that these are the only nonplanar graphs!

@@

\advsect Basic counting techniques

@@

\medskip\boldlabel Counting strings.
Back in the proof of Proposition~{\propcardpowerset}, we counted the number of binary strings of
length $n$ to be $2^n$. More generally, if there are $s$ different symbols and $t$ different
slots in which to put them, then the number of strings is $s^t$. This can be phrased
in terms of the number of functions from a set $T$ of size $t$ to a set $S$ of size $s$.
Indeed, people often use the notation $S^T$ to denote the set of all functions from $T$ to $S$,
and the notation should remind you that $|S^T| = |S|^{|T|}$.

@@

\medskip\boldlabel The principle of inclusion and exclusion. As our next toy problem,
suppose we want to count the number of strings of length $6$, using the symbols
$a$, $b$, $c$, and $d$, that start with $a$ or end with $aa$ (or both). The number
of strings that start with $a$ is $4^5$, since we have no choice for the first symbol,
but $4$ choices for each of the remaining five symbols; likewise, the number
of strings that end with $aa$ is $4^4$. Adding these two up, we get $4^5+4^4$.
There's a problem though. We have accidentally double-counted the strings that
both start with $a$ {\it and} end with $aa$. How many of these are there? Well, the first, fifth, and
sixth slots all have to be $a$, but there are four choices for each of the remaining
three slots, for a total of $4^3$. Subtracting this from our subtotal yields
the correct answer: $4^5+4^4-4^3$.
\endgroup%\headline{\hfil}\footline{\hfil}

This technique works with any two sets $A$ and $B$,
i.e., $|A\cup B| = |A| + |B| - |A\cap B|$. For three sets $A$, $B$, and $C$,
we have
$$|A\cup B\cup C| = |A| + |B| + |C| - |A\cap B| - |A\cap C| - |B\cap C| + |A\cap B\cap C|$$
(draw a Venn diagram and convince yourself that this is correct).
For a general number of sets, we have the following theorem.

\proclaim Theorem {\advthm} (Principle of inclusion and exclusion).

>>>>

\vskip36pt

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
graphs}{74}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 15.\kern .5em Eulerian
trails and circuits}{81}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx IV. Combinatorics \kern 3pt}}{88}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 17.\kern .5em Basic counting techniques}{89}}

@@

\medskip\boldlabel\datestamp{10}{ix} Conditional and biconditional. We now examine the {\it conditional}
logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\To q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 1 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}.
$$
Within the conditional statement, $p$ is called the {\it antecedent}; this is the assumption. The
statement that is asserted, conditional that the antecendent holds, is called the {\it consequent} $q$.
You can quickly check that the relation $p\To q$ is equivalent to $\neg p \vee q$. This fact is useful
when performing mechanical simplifications.
In English, the statement ``if $p$ then $q$'' asserts a causal relation between $p$ and $q$.
Take a moment and reconcile this idea with the truth table above.
It is normal to get a little tripped up if it's your first time seeing this table.
In the first row, $q$ doesn't even happen, so it might feel weird to say that $p$
has ``caused'' $q$ to happen in this case, and
in the second row, $p$ is not true
and $q$ is true, so it might seem odd that we have set $p\To q$ to true, because there seems to be no relation
between $p$ and $q$.

But it should make sense if you think
of $p$ as a precondition to a promise $q$, and then consider
whether the promise is broken. As an example,
suppose your friend says, If it snows tomorrow I'll work in Trottier with you. If it doesn't snow
and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
snow and she shows up, then she still hasn't broken her promise. The only way she can break her promise is if
it snows and she doesn't come to Trottier; this situation corresponds to the only $0$-row in the truth table.

@@

\edef\propoddevensubsets{\the\thmcount}
\proclaim Proposition \advthm. Let $X$ be a finite nonempty set.
Let $E$ be the set of all subsets of $X$ with
even cardinality, and let $D$ be the set of all subsets of $X$ with odd cardinality.
Then $|E| = |D|$.

@@

\edef\thmpnt{\the\thmcount}
\parenproclaim Theorem {\advthm} (Prime number theorem). For $x\in \RR$, let
$$\pi(x) = \bigl| \{ p \le x : p\ \hbox{prime}\} \bigr|.$$
Then $\pi(x) \sim x/\ln x$, in the sense that
$$\lim_{x\to\infty} {\pi(x)\over x/\ln x} = 1.$$

@@

\noindent\datestamp{19}{xi} \hskip\parindent
Not only are all graphs that have $K_5$ or $K_{3,3}$ as a minor nonplanar, it turns out
that these are the only nonplanar graphs!

@@

\advsect Counting techniques

@@

\medskip\boldlabel Counting strings.
Back in the proof of Proposition~{\propcardpowerset}, we counted the number of binary strings of
length $n$ to be $2^n$. More generally, if there are $s$ different symbols and $t$ different
slots in which to put them, then the number of strings is $s^t$. This can be phrased
in terms of the number of functions from a set $T$ of size $t$ to a set $S$ of size $s$.
Indeed, people often use the notation $S^T$ to denote the set of all functions from $T$ to $S$,
and the notation should remind you that $|S^T| = |S|^{|T|}$. Identifying the number
$2$ with the set $\{0,1\}$, the power set $2^X$ is in bijection with the set $\{0,1\}^X$, so
this notation for the set of all functions is consistent with our earlier notation for
power set.

@@

\medskip\boldlabel The principle of inclusion and exclusion. As our next toy problem,
suppose we want to count the number of strings of length $6$, using the symbols
$a$, $b$, $c$, and $d$, that start with $a$ or end with $aa$ (or both). The number
of strings that start with $a$ is $4^5$, since we have no choice for the first symbol,
but $4$ choices for each of the remaining five symbols; likewise, the number
of strings that end with $aa$ is $4^4$. Adding these two up, we get $4^5+4^4$.
There's a problem though. We have accidentally double-counted the strings that
both start with $a$ {\it and} end with $aa$. How many of these are there? Well, the first, fifth, and
sixth slots all have to be $a$, but there are four choices for each of the remaining
three slots, for a total of $4^3$. Subtracting this from our subtotal yields
the correct answer: $4^5+4^4-4^3$.

This technique works with any two sets $A$ and $B$,
i.e., $|A\cup B| = |A| + |B| - |A\cap B|$. For three sets $A$, $B$, and $C$,
we have
$$|A\cup B\cup C| = |A| + |B| + |C| - |A\cap B| - |A\cap C| - |B\cap C| + |A\cap B\cap C|$$
(draw a Venn diagram and convince yourself that this is correct).
For a general number of sets, we have the following theorem.
\endgroup%\headline{\hfil}\footline{\hfil}

\edef\thminclusionexclusion{\the\thmcount}
\parenproclaim Theorem {\advthm} (Principle of inclusion and exclusion). Let $n\ge 2$
be an integer and let $A_1,\ldots, A_n$ be finite sets. Then
$$\biggl| \bigcup_{i=1}^n A_i\biggr|
=\sum_{k=1}^n (-1)^{k+1}\biggl(\sum_{1\le i_1<\cdots<i_k\le n}|A_{i_1}\cap\cdots\cap A_{i_k}|\biggr).$$

This sum of sums might look intimidating, but breaking down what it says bit by bit,
we see that it is simply a sum over all possible nonempty intersections of the $A_i$, with a sign
according to how many of the $A_i$ we are intersecting. We will prove this theorem soon,
after we state and prove the binomial theorem. We first define, for nonnegative integers
$n$ and $k$, the {\it binomial coefficient} ${n\choose k}$ to be the number of ways of choosing
$k$ distinct elements out of a pool of $n$ distinct elements. (The quantity ${n\choose k}$
is read ``$n$ choose $k$.'') We have the formula
$${n\choose k} = {n!\over k!(n-k)!} = {n(n-1)\cdots (n-k+1)\over k(k-1)\cdots 1}.$$
This is because we can pick $k$ elements out of an $n$-element set by permuting all the $n$ elements
(there are $n!$ ways to do this), and then simply looking at the first $k$ elements. But the order
of the $k$ elements we choose doesn't matter, so we arrive at the same elements no matter which
of the $k!$ possible orders they might appear in. Same goes for the $(n-k)!$ orders the elements
that we {\it didn't} pick can appear in.

The name ``binomial coefficient'' is justified by the following theorem, first stated and
proved by Al-Karaji in around the tenth century.

\parenproclaim Theorem {\advthm} (Binomial theorem). For all $x,y\in \RR$ and positive integers $n$,
$$(x+y)^n = \sum_{k=0}^n {n\choose k} x^k y^{n-k}.$$

\proof Expanding the left-hand side
$$(x+y)^n = (x+y)(x+y)\cdots (x+y),$$
we see that there will be $2^n$ different terms after repeatedly using the distributive law.
Each term will be a product of $x$ to some power and $y$ to some power, where the powers add up to $n$.
In other words, each term looks like $x^ky^{n-k}$ for some $0\le k\le n$. The number of times
that the term $x^k y^{n-k}$ appears in the expansion
is the number of ways to choose $k$ items out of a pool of $n$, namely,
${n\choose k}$. This gives exactly the right-hand side.\slug

We are now able to prove the principle of inclusion and exclusion.

\medskip\noindent{\it Proof of Theorem~\thminclusionexclusion.}\enspace
Let $n\ge 2$ and let $A_1,\ldots,A_n$ be finite sets.
Recall that the identity we want to prove is
$$\biggl| \bigcup_{i=1}^n A_i\biggr|
=\sum_{k=1}^n (-1)^{k+1}\biggl(\sum_{1\le i_1<\cdots<i_k\le n}|A_{i_1}\cap\cdots\cap A_{i_k}|\biggr).$$
Let $x$ be an element that belongs to the union $\bigcup_{i=1}^n A_i$. Clearly, this contributes
$1$ to the left-hand side, so we must show that it contributes exactly $1$ to the right-hand side
as well. We know that $x$ is a member of at least one of the $A_i$;
let $s\ge 1$ be the number of sets $A_i$ that contain $x$.

Each time $x$ is a member of
some $A_i$, it contributes $+1$ to the right-hand side. This happens ${s\choose 1}$ times.
Each time it is a member of $A_i$ and
$A_j$ for $i<j$, $x$ contributes $-1$ to the-right hand side; there are ${s\choose 2}$ of these terms.
Generally, if $x$ is a member
of $A_1\cap A_2\cap\cdots\cap A_r$, it contributes $+1$ if $r$ is odd and $-1$ if $r$ is even,
and this term is repeated ${s\choose r}$ times. Hence $x$ contributes
$${s\choose 1} - {s\choose 2} + {s\choose 3} - \cdots + (-1)^{s+1} {s\choose s}$$
to the right-hand side.

But, by the binomial theorem,
$$0 = (1-1)^s = {s\choose 0} - {s\choose 1} + {s\choose 2} -\cdots + (-1)^s{s\choose s}.$$
We know that ${s\choose 0} = s!/(0! s!) = 1$, so by rearranging we find that
$${s\choose 1} - {s\choose 2} + {s\choose 3} - \cdots + (-1)^{s+1} {s\choose s} = {s\choose 0} = 1,$$
which means that the contribution of $x$ to the right-hand side is $1$.
We are done since $x$ is arbitrary.\slug

As an application of this principle, let $X = \{1,2,\ldots, 100\}$ and suppose we want
to count the number of $n\in X$
with $\gcd(n,30) = 1$. To accomplish this, it turns out to be easier to count the number
of $n\in X$ that have $\gcd(n,30)\ge 2$ (and then we must subtract this number from $|X| = 100$).
For any positive integer $r$, let
$$A_r = \{n\in X : r\divides n\}.$$
Since $30 = 2\cdot 3\cdot 5$, an integer $n\in X$ has $\gcd(n,30)\ge 2$ if and only if
$n\in A_2$, $n\in A_3$, or $n\in A_5$.
By the principle of inclusion and exclusion,
$$|A_2\cup A_3\cup A_5| = |A_2| + |A_3| + |A_5|
- |A_2\cap A_3| - |A_2\cap A_5| - |A_3\cap A_5| + |A_2\cap A_3\cap A_5|.$$
The number of even integers in $X$ is $100/2 = 50$; similarly, the number
of integers divisible by $3$ in $X$ is $\lfloor 100/3\rfloor = 33$. In general,
$|A_r| = \lfloor 100/r\rfloor$. Furthermore, since $\gcd(2,3) = 1$,
the intersection $A_2\cap A_3$ is simply $A_6$, and by analogous reasoning, we see that
$$\eqalign{
|A_2\cup A&_3\cup A_5| \cr
&= |A_2| + |A_3| + |A_5| - |A_6| - |A_{10}| - |A_{15}| + |A_{30}| \cr
&= \biggl\lfloor {100\over 2}\biggr\rfloor
+\biggl\lfloor {100\over 3}\biggr\rfloor
+\biggl\lfloor {100\over 5}\biggr\rfloor
-\biggl\lfloor {100\over 6}\biggr\rfloor
-\biggl\lfloor {100\over 10}\biggr\rfloor
-\biggl\lfloor {100\over 15}\biggr\rfloor
+\biggl\lfloor {100\over 30}\biggr\rfloor\cr
&= 50 + 33 + 20 - 16 - 10 - 6 + 3 \cr
&= 74.
}$$
We conclude that the number of $n\in X$ such that $\gcd(n,30) = 1$ is
$100-74 = 26$.

\advsect Permutations and combinations

Suppose we have a pool of $100$ hockey players and we want to dress $20$
of them for a game, giving them the numbers $1$ through $20$. How many ways are there to do this?
Well, there are $100$ choices for the player with jersey number $1$, then after
picking that player, there are $99$ choices for the player with jersey number $2$, and so on,
so the total number is
$$100\cdot 99\cdot 98 \cdots 82\cdot 81 = {100!\over 80!}.$$
\bye

\noindent\datestamp{23}{} \hskip\parindent
In general, we define a {\it $k$-permutation} of a set $X$ to be an ordered subset
of $X$ (i.e., the order of the selection matters) of size $k$. If $|X| = n$,
the number of $k$-permutations of $X$ is denoted $P(n,k)$ and equals
$$ P(n,k) = n\cdot (n-1)\cdots (n-k+1) = {n!\over (n-k)!}.$$
This number is also sometimes called the {\it falling factorial} and denoted $n^{\underline k}$.
It is the same as the number of functions from $\{1,2\ldots, k\}$ to $X$ that are injective.
(Pause a moment to think about it if you need to.) Recall that if $|S| = s$ and $|T| = t$,
the total cardinality of $S^T$ is $s^t$. Well, the number of
{\it injective} functions from $T$ to $S$ is $P(s,t) = s^{\underline{t}}$.

An $n$-permutation of a set of size $n$ is also just called a {\it permutation} of
$n$. There are
$$P(n,n) = {n!\over (n-n)!} = n!$$
of these. (Note that $0! = 1$, by definition.)

To illustrate various counting techniques associated with permutations, consider the following
situation. We have ten books: three comic books, three books of poetry, and four novels.
How many ways are there to order all of them on a shelf? The answer is the number of permutations
on $10$ objects, namely, $10! = 3628800$. How about just putting five of them on the shelf?
Then the answer
is $P(10,5) = 10!/5! = 30240$. Lastly, what if we want to put all of them on the shelf,
but group all the comic books together, all the poetry books together, and all the novels together.
There are $3!$ ways to order the comic books amongst themselves, $3!$ for the poetry books
as well, and then $4!$ for the novels. Then we must arrange the ordered bundles of books
on the shelf; there are three bundles so there are $3!$ ways to do this. Thus the total
number of ways is $3!3!4!3! = 5184$.

Returning to the hockey-team example, suppose we still want to pick a team of $20$ out of
$100$ possible players, but now we don't number the players; that is, we only care about
the sets of the players themselves, not the ordering amongst them. The number $P(100,20)$
is $20!$ times too big, since there are $20!$ different orderings of the team we choose,
and each of them is only counted once in our new scenario. So the answer is
$${P(100,20)\over 20!} = {100!\over 20!(100-20)!} = {100\choose 20}.$$
We already had an intimate encounter with
the quantity ${n\choose k}$ in the previous section. Defining a
{\it $k$-combination} of a set $X$ to be the number of ways of choosing $k$ elements of of
$X$, the number of $k$-combinations of a set of size $n$ is ${n\choose k}$.

\medskip\boldlabel Counting grid paths.
An application of $k$-combinations is counting the number of paths in a grid.
If you have ever had the pleasure or misfortune of visiting
Edmonton, Alberta, you will know that most of the city's roads
are organised in a numbered grid pattern. Avenues run east to west and streets run north to south.
Suppose we have just enjoyed a cortado in the trendy caf\'e
at the corner of 82nd Avenue and 104th Street,
and now we want to get to the sandwich shop at the corner of 85th Avenue and 109th Street.
To do so we will have to walk three blocks north and five blocks west. How many ways are there
to do this?
\midinsert
$$\epsfbox{edmonton.ps}$$
\vskip5pt
\caption{One possible path from the caf\'e to the sandwich shop.}
\endinsert
\goodbreak
Well, the journey is eight blocks in length, so each possible path can be expressed as
a string of eight symbols, call them {\mc N} for north and {\mc W} for west. We also know there
must be exactly three {\mc N}s and five {\mc W}s; for example, one possible path, pictured in
Fig.~12, is {\mc NWWWNWWN}. There are ${8\choose 3}$ different ways to choose where the {\mc N}s
go in the string, after which the positions of the {\mc W}s is completely determined. Alternatively,
we can pick where the {\mc W}s go, in ${8\choose 5}$ ways. Thus the number of possible paths
from the caf\'e to the sandwich shop is
$${8\choose 3} = {8!\over 3! 5!} = {8\choose 5} = 56.$$
Here we see an important principle in action. {\sl The number of $k$-combinations of a set
of size $n$ is the same as the number of $(n-k)$-combinations of the same set.}
This is because choosing $k$ items is
the same as choosing which $n-k$ items {\it not} to take.
In general, the number of ways of walking $m$ blocks in one direction and $n$ blocks in a perpendicular
direction is
$${m+n\choose m} = {m+n\choose n}.$$

\medskip\boldlabel Stars and bars.
Suppose there is a shop selling marbles that come in six colours. How many ways are there to select
a bag of four marbles? Imagine that the marbles are laid out in six different bins all in a row,
one for each colour. We stand at the first bin and may take some number of them. Mark down a
star for each one that we take. Then we step forward to the next bin; record this action by
drawing a bar. If we take some marbles here, mark down a star for each marble taken. Then we step forward
again, so mark down a bar. At the end, we must have taken exactly four marbles and switched bins
five times, so there must be exactly four stars and five bars, e.g., the sequence
$$\mid\quad\mid\star\mid\star\mid\quad\mid\star\star$$
corresponds to taking one marble of the third colour, one of the fourth, and two of the sixth.
There are nine slots, four of which have to be stars, so
the number of ways to pick four marbles out of six bins is ${9\choose 4} = 126$. In general
the number of ways to pick $m$ marbles out of $b$ bins is ${m+b-1\choose m}$. This method of
counting is called the {\it stars-and-bars} technique.

\medskip\boldlabel A night at the casino.
Suppose we are playing a high-stakes card game at the local gambling
house and it is somehow pertinent for us to calculate how many five-card hands there
are with at least two hearts. A na\"\i ve attempt would run as follows. We pick two cards
to be hearts, in ${13\choose 2}$ ways, then there are $50$ cards left to choose from,
and we can pick any three of those, in ${50\choose 3}$ ways. This seems logical, except
that we have counted some hands twice! For instance, suppose that in the first step we pick
our two hearts to be $\tt2\heartsuit$ and $\tt6\heartsuit$, then for our remaining three
cards we pick $\tt8\heartsuit$, $\tt3\diamondsuit$, and $\tt Q\spadesuit$. We can get the exact
same hand by first picking our two hearts to be $\tt6\heartsuit$ and $\tt8\heartsuit$,
and then choosing $\tt2\heartsuit$, $\tt3\diamondsuit$,
and $\tt Q\spadesuit$ for the remaining three cards.

One correct way of getting the correct count is to count the number of hands with
exactly two hearts, exactly three hearts, exactly four hearts, and exactly five hearts. Since
these possibilities all define disjoint sets, we can simply add them up at the end.
The number of ways of drawing exactly two hearts is
$${13\choose 2}{52-13\choose 3} = {13\choose 2}{39\choose 3} = 712842,$$
since there are $39$ non-heart cards and we must select three of them.
Similarly, the number of ways of drawing exactly three hearts is
$${13\choose 3}{52-13\choose 2} = {13\choose 3}{39\choose 2} = 211926;$$
we also calculate
$${13\choose 4}{39\choose 1} = 27885\qquad\hbox{and}\qquad
{13\choose 5}{30\choose 0} = 1287.$$
Hence the number of five-card hands with at least two hearts is
$$712842 + 211926 + 27885 + 1287 = 953940.$$
This number has little practical value unless we know the number of five-card hands
in total; this is ${52\choose 5} = 2598960$. We can then calculate the probability
of a five-card hand containing at least two hearts to be $953940/2598960 \approx 36.7\%$.

Deriving the total number of five-card hands suggests a different way to get
the same count.
We could take this total and subtract the number
of hands with no hearts, as well as the number of hands with exactly one heart. This gives
$${52\choose 5} - {13\choose 0}{39\choose 5} - {13\choose 1}{39\choose 4}
= 2598960 - 575757 - 1069263 = 953940,$$
the same figure we obtained in the previous paragraph.

\medskip\boldlabel\datestamp{24}{} Identities involving binomial coefficients.
We already observed earlier that the number of ways of picking $k$ objects out of a
set of size $n$ is the same as the number of picking $n-k$ objects out of the same set.
This can be seen algebraically by expanding
$${n\choose k} = {n!\over k!(n-k)!} = {n!\over (n-k)!\bigl(n-(n-k)\bigr)!} = {n\choose n-k}.$$
Here is another identity involving binomial coefficients.

\proclaim Proposition \advthm. Let $n\ge 0$ be an integer. Then
$$\sum_{k=0}^n {n\choose k}^2 = {2n\choose n}.$$

\proof Let $X$ be set of size $2n$. The right-hand side counts the number of subsets of
$X$ of size $n$. We count this in a different way. Let $A$ and $B$ be such that $|A| =|B| = n$,
and $A\cup B = X$ (so we must have $A\cap B = \emptyset$). Then choosing a subset
of $X$ is the same as choosing $k$ elements of $A$, where $0\le k\le n$,
and then choosing the remaining $n-k$ elements from $B$. In other words,
$${2n\choose n} = \sum_{k=0}^n {n\choose k}{n\choose n-k}
= {n\choose k}^2.\noskipslug$$

More identities can be obtained by laying out binomial coefficients in the triangle
$$\matrix{
&&&{0\choose 0}&&&\cr
&&{1\choose 0}&&{1\choose 1}&&\cr
&{2\choose 0} && {2\choose 1} && {2\choose 2}&\cr
{3\choose 0} && {3\choose 1} && {3\choose 2} && {3\choose 3} \cr
\vdots && \vdots && \vdots && \vdots && \cr
}$$
Adding a few more rows and writing out the actual values of the binomial coefficients
yields Fig.~13. This triangle is named after B.~Pascal, who wrote a treatise regarding some of
its properties that was published posthumously in 1665. (The triangle is known by different
names in some different countries, as it was studied by Al-Karaji in the tenth or
eleventh century, by O.~Khayy\'am about a hundred years later, by Yang Hui in
the thirteenth century, and by Tartaglia in 1556, among many others.)
\midinsert
$$\matrix{
&&&&&&&& 1 &&\cr
&&&&&&& 1 && 1 &&\cr
&&&&&& 1 && 2 && 1 &\cr
&&&&& 1 && 3 && 3  && 1 \cr
&&&& 1 && 4 && 6 && 4 && 1 \cr
&&& 1 && 5 && 10 && 10 && 5 && 1 \cr
&& 1 && 6 && 15 && 20 && 15 && 6 && 1 \cr
& 1 && 7 && 21 && 35 && 35 && 21 && 7 && 1 \cr
1 && 8 && 28 && 56 && 70 && 56 && 28 && 8 && 1 \cr
\vdots &&\vdots && \vdots && \vdots && \vdots && \vdots && \vdots && \vdots &&\vdots \cr
}$$
\vskip5pt
\caption{The first nine rows of Pascal's triangle}
\endinsert
\goodbreak
This triangle can easily be drawn by hand by writing $1$s along its boundary, and then filling it
it using the observation that every number is the sum of the two directly above it in the previous
row. Of course, this rule has to be proved, so we shall do so right now.

\parenproclaim Theorem {\advthm} (Pascal's identity). Let $n\ge 0$ and $k\ge 1$ be integers.
Then
$${n\choose k} = {n-1\choose k} + {n-1\choose k-1}.$$

\proof Set $X = \{1,2,\ldots,n\}$. We count the number of sets $Y\subseteq X$ there are
of size $k$.
Either $1\notin Y$ or $1\in Y$. If $1\notin Y$, then there are ${n-1\choose k}$ possibilities
for what $Y$ can be. If $1\in Y$, then there are still $k-1$ other elements of $Y$, and $n-1$
elements to choose from, so there are ${n-1\choose k-1}$ possiblities in this case. Hence
the right-hand side counts the number of subsets of $X$ with cardinality $k$. But this is
exactly what the left-hand side counts as well.\slug

Another thing we notice about Pascal's triangle is that the first row sums to $1$, the second
row sums to $2$, the third to $4$, the fourth to $8$, and so on. We might therefore
conjecture that the $n$th row of Pascal's triangle should always sum up to $2^n$. Indeed, we
have the following easy corollary of the binomial theorem.

\proclaim Proposition \advthm. For all $n\in \NN$,
$$\sum_{k=0}^n {n\choose k} = 2^n.$$

\proof By the binomial theorem with $x=y=1$, we have
$$2^n = (1+1)^n = \sum_{k=0}^n {n\choose k} 1^k 1^{n-k} = \sum_{k=0}^n {n\choose k}.$$
The proposition can also be proven by noting that the left-hand side counts the number
of subsets of $\{1,2,\ldots,n\}$ of size $0$, plus the number of subsets of size $1$,
and so on up to all subsets of size $n$. But adding these all up simply gives the total number
of subsets of $\{1,2,\ldots, n\}$, which we know to be $2^n$.\slug

The binomial theorem with a different choice of $x$ and $y$ allows us to reprove a statement
we showed a long time ago, that for any finite nonempty set $X$, the number of subsets of $X$ with
odd cardinality is the same as the number of subsets of $X$ with even cardinality.

\medskip\noindent{\it Another proof of Proposition~{\propoddevensubsets}}.\enspace
We re-establish the notation of the proposition statement. Let $X$ be a finite nonempty set,
let $E$ be the set of subsets of $X$ with even cardinality, and let $O$ be the set of
subsets with odd cardinality. By the binomial theorem with $x=-1$ and $y = 1$, we have
$$0 = \bigl(-1+1\bigr)^n = \sum_{k=0}^n {n\choose k} (-1)^k 1^{n-k}
= {n\choose 0} - {n\choose 1} + {n\choose 2} -\cdots + (-1)^n{n\choose n}.$$
Hence
$${n\choose 0} + {n\choose 2} +\cdots = {n\choose 1} + {n\choose 3} + \cdots,$$
but the left-hand side of this is just $|E|$, and the right-hand side is $|O|$.\slug
